How predictable is extinction? Forecasting species survival at million-year timescales

### 2019-04-26 doc/paper.tex
seth 4th rnd comments in

Extinction intensity (average rate) and selectivity (difference in risk between taxa) vary through time and the relative risk of extinction exhibited by different taxonomic groups and how that risk varies over time is an important dynamic which shapes the rate and structure of extinction \citep{Payne2007,Payne2016,Ezard2011,Smits2019}. What has not been evaluated is that as extinction intensity and selectivity change over time, how accurate are our assessments based on past events likely to be when applied to the future? Given this variation, how accurate are our assessments based on past events likely to be when applied to the future? Putting aside the important question of how human activities will alter the determinants of future extinction risk, we can address this uncertainty by specifically including and modeling the temporal variation in extinction risk across a range of extinction intensities and selectivities.

Numerous studies have established that geographic range is one of the most important determinants of extinction risk in the fossil record, and that a species geographic range can be highly variable over geologic time \citep{Foote2007,Liow2010,Liow2007,Kiessling2013,Payne2007,Jablonski2003,Jablonski2008,Jablonski2006}. The degree to which the past can help to predict the future fates of species depends in part on the degree to which species’ geographic range trajectories are deterministic versus Markovian. In the former case, knowledge of the specific past trajectory of a species -- whether its range has expanded or contracted over a given time span -- might help to improve assessments of its current risk. In the latter case, only the current range of the species conveys useful information about its current risk, although we can still use prior extinction patterns to augment predictions by evaluating the relative extinction risk of species that had similar ranges in the past. Discriminating among these alternative models is thus very important for determining how best to incorporate fossil data in present risk assessments.

@@

We analyzed microfossil occurrence information from the Neptune Database \url{http://www.nsb-mfn-berlin.de/nannotax} \citep{Lazarus1994,SpencerCervato1999}. This dataset of occurrences covered calcareous nannoplankton, diatoms, foraminifera, and radiolarians.% -- these occurrences span the entire globe between 120 and 0 million years ago (Mya). 
These occurrences were then filtered to include only those species who's first occurrence was no more than 63 Mya. This filtering criterion means that our analysis avoids those taxa which survived the K/Pg boundary, those taxa which arose just after the K/Pg boundary, and ensures that our occurrence histories are congruent with the temperature time-series used as a predictor of extinction (see below).

@@

Average global temperature of each 1 My bin was calculated from estimates based on Magnesium/Calcium isotope ratios \citet{Cramer2011}. Elemental ratios are considered more accurate estimates of past global temperature when compared to the frequently used Oxygen isotope-based estimates; this is because Mg/Ca based estimates are not effected by ice-volume and fresh-water input (e.g. meteoric water) which can alter Oxygen isotope ratios without reflecting changes to the climate itself. This property is of particular importance for this analysis as polar ice-caps develop midway through the Cenozoic. Our data source, \citet{Cramer2011}, estimated temperature for every 0.1 My from 0 to 63 Mya. The temperature estimate for each 1 My interval was calculated as the mean of all estimates within that interval.

@@

We used a discrete-time survival modelling framework in order to estimate how well we can predict extinction risk at one million year time scales. At its core, our model is a multilevel logistic regression with taxon age in millions of years as a varying intercept \citep{Tutz2016}. We considered four variations of the predictors included in our model: covariate effects are constant over time and none of our historical covariates are included (Model C), covariate effects are allowed to vary over time but none of our historical covariates are included (Model V), covariate effects are constant over time and our historical covariates are included (Model CP), covariate effects are allowed to vary over time and historical covariates are included (Model VP). We always included species geologic age at time of observation as a non-nested varying intercept term. This factor may or may not contribute to differences in species extinction risk over time \citep{Smits2015,Finnegan2008,Ezard2012,VanValen1973,Liow2011,Crampton2016}, but its inclusion in our model is critical to its nature as a survival model \cite{Tutz2016}.

@@

We are interested in our models' performance into two contexts: in-sample and out-of-sample predictive performance i.e. forecasting.

@@

We are particularly interested in understanding how well our model forecasts extinction probability of data from the future that the model was not fit to (out-of-sample data). To quantify our ability to forecast species' extinction risk, we estimated average out-of-sample forecasting performance using 5-fold time-series cross-validation. For time-series data, the folds (data partitions) are approximately equal segments of time. Each fold represents a sequence of time points. With 63 time points, each of the five folds represents approximately 13 time points. Keep in mind, however, that each time point corresponds to many (100-1000) individual observations.

@@

The in-sample model comparisons are useful comparing the relative ability of our models to represent the data it was fit to. Comparison between the posterior distributions of in-sample AUC for each of the four models demonstrates that all of our models have approximately equal in-sample forecasting performance (Fig. \ref{fig:auc_hist}). The parameter rich model VP has the greatest median in-sample AUC when compared to the other three models, but there is substantial overlap in their posterior distributions. Additionally, while our parameter rich model VP is possibly the most adequately performing model, the difference or improvement to performance is minimal at best -- all four models have approximately equal in-sample AUC posterior distributions. All the in-sample AUC estimates from our models are concentrated on an AUC value of 0.77. It is therefore hard to conclude that there is one ``best'' model which we can rely upon as they are all nearly functional equivalent.

@@

There are one or more time periods where our posterior distribution of in-sample AUC has a median value less than or equal to 0.5 (Fig. \ref{fig:auc_taxon_time}). However, this pattern is absent for the posterior estimates of in-sample AUC for Foraminfera and Radiolaria as fit by the VP model. In contrast, there are fewer periods of low model performance for calcareous nannoplankton and Dinoflagellates as estimated from our VP model than in those estimates from the other three model variations.

@@

Expected out-of-sample forecasting performance was estimated using five-fold cross-validation for time series \citep{Arlot2010,Bergmeir2016}. The resulting distribution when all folds are combined is very multimodal by their very nature of being fit to and estimated from different data sets and amounts of data \citep{ESL}. We find that this multimodality increases with model complexity (Fig. \ref{fig:fold_auc}), most likely because the complex models allow for predictor effects to vary with time which allows for a greater range in possible parameter values which in turn yield a greater range of posterior predictions.

Comparison between the in-sample forecasting performance estimates to the expected out-of-sample AUC estimates reveals a similar range in performance for all models (Fig. \ref{fig:auc_hist}, \ref{fig:fold_auc}). Interestingly, the differences in posterior predictive distributions of AUC between the four model variations has decreased. For example, the VP model no longer has the greatest median AUC of the four models (Fig. \ref{fig:fold_auc}). These differences mean that the rank order of median out-of-sample AUC is different from the rank order of median in-sample AUC. However, making interpretations bsaed only incorrect -- our estimates are the full posterior, not individual points. Because of this, the four models are effectively indistinguishable in their out-of-sample forecast performance (Fig. \ref{fig:fold_auc}).

@@

We can also compare the posterior predictive distribution of expected out-of-sample AUC over time and taxonomic group for each of the four models (Fig. \ref{fig:fold_auc_taxon_time}). In all cases, our four models are nearly identical in their out-of-sample forecasting performance over time and between each of the taxonomic groups.

In the analysis of the in-sample forecast performance of the four models, we noted that there were time intervals where our predictions were no better than random (Fig. \ref{fig:auc_taxon_time}). This occurrence is generally much rarer for the posterior distribution of AUC from the out-of-sample forecasts. The major exception to this pattern are our estimates for the Dinoflagellates, which has at least one time interval for all four models where the median AUC of the out-of-sample forecasts were no better random. In contrast, the only other group for which median posterior predictive estimate of out-of-sample AUC reaches 0.5 is calcareous nannoplankton, and then only with model V.
\begin{figure}[ht]
 \centering
 \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_taxon_time_full}
 \caption{Comparison of out-of-sample AUC values over time as aggregated by taxonomic group for each of the four models. The AUC of the individual My intervals within each fold is plotted to highlight the heterogeneity in performance within and between folds. This presentation decomposes each of the 12-million year folds by each of the taxonomic groups into the predictions made for each of the million-year intervals. The black line corresponds to the median AUC estimate, with the envelopes corresponding to multiple credible intervals as indicated in the legend. See Table \ref{tab:model_def} for a description of each of the four models.}
 \label{fig:fold_auc_taxon_time}
\end{figure}

We compared the difference in our AUC estimates from the out-of-sample forecasts to the AUC estimates from our in-sample forecasts by subtracting the in-sample AUC estimates from the out-of-sample AUC estimates (Fig. \ref{fig:oos_ins_diff}). A difference in AUC close to 0 indicates complete congruence between the in-sample and out-of-sample forecasts. A positive difference indicates that our out-of-sample forecasts are actually higher performing than our in-sample forecasts, while negative difference indicates poorer out-of-sample performance than in-sample forecast. Divergences between our out-of-sample and in-sample forecasts are rare and tend to not form multimillion year patters, something that is consistent with the broad visual congruence between the in-sample and out-of-sample forecast performance (Fig. \ref{fig:auc_taxon_time}, \ref{fig:fold_auc_taxon_time}). The only major multimillion year pattern indicating significantly poorer out-of-sample forecast performance than in-sample forecast performance is for Radiolaria based on the VP model concentrated around 30 Mya (Fig. \ref{fig:oos_ins_diff}).

@@

We find that all of our models are expected to correctly forecasting the rank order in extinction probabilities for a randomly selected extinct-extant pair of future observations between 70\% to 80\% of the time (Fig. \ref{fig:fold_auc}). One of the most striking aspects of these results is the similarity in forecasting performance for in-sample and out-of-sample observations. A slight decrease in performance when dealing with out-of-sample observations makes sense: each of the models fit during cross-validation is based on fewer data than the model fit on the full data (between 1/5th to 4/5ths of the original). Additionally, a potential decrease in precision when forecasting the future of extinction risk is almost to be expected as the future is not necessarily the same as the past. However, the similarly between the in-sample and out-of-sample results indicates that our model is fairly robust to how extinction intensity has changed over the Cenozoic. An important caveat to our results is that human impacts may substantially alter extinction risk dynamics away from the average Cenozoic ``state'' so that the future could become less predictable than it has been in the past.

@@

We also find that our four models are practically identical in their ability to make in-sample and out-of-sample forecasts. Although the in-sample AUC estimates differ between models, all of these estimates are in a narrow range of possible AUC values (Fig. \ref{fig:auc_hist}). While the ``best'' model does include the historical covariates and allows all covariate effects to vary over time, its practical difference in performance versus the other models is negligible (Fig. \ref{fig:fold_auc}). This means that even though model VP has a statistically greater AUC for in-sample forecasts than the other three models, this result is not practically or scientifically significant. The results of the out-of-sample forecasts illustrate this point in full, as all four models have functionally identical out-of-sample performance.

@@

The quality and consistency of our models out-of-sample forecasting performance is surprising given that its estimates are based on rather limited biological and environmental information about the taxa. Our most complex models are only taking into account some aspects of species geographic range and their taxonomic grouping. The principal reason we were not able to include more biological information in our models is that we lack almost any life history or ecological information on most marine micro- and nannoplankton. Forams are an exception to this problem as aspects of life history, ecology, and physiology for many foram species \citep{Ezard2011}. However, comparable information does not exist all foram species, nor does this type of data exist for the other three taxonomic groups studied here. Including this type of information in a predictive model of extinction risk we require analyzing a single, limited taxonomic subset of the fossil occurrences present the Neptune Database. This would mean ignoring the majority of occurrence information present in the Neptune database.

@@

In conclusion, our results provide a promising picture of our ability to predict the relative extinction risk of two randomly selected species. Considering that conservation decisions are made based on a continuum of risk, from most to least, this means that our results are in the same language as how conservation resources are allocated.

>>>>

Extinction intensity (average rate) and selectivity (difference in risk between taxa) vary through time and the relative risk of extinction exhibited by different taxonomic groups and how that risk varies over time is an important dynamic which shapes the rate and structure of extinction \citep{Payne2007,Payne2016,Ezard2011,Smits2019}. This variation also raises a question: given that extinction intensity and selectivity change over time, how accurate are our assessments based on past events likely to be when applied to the future? Putting aside the important question of how human activities will alter the determinants of future extinction risk, we can address this uncertainty by specifically including and modeling the temporal variation in extinction risk across a range of extinction intensities and selectivities.

Numerous studies have established that geographic range is one of the most important determinants of extinction risk in the fossil record, and that a species geographic range can be highly variable over geologic time \citep{Foote2007,Liow2010,Liow2007,Kiessling2013,Payne2007,Jablonski2003,Jablonski2008,Jablonski2006}. The degree to which the past can help to predict the future fates of species depends in part on the degree to which species’ geographic range trajectories are deterministic versus Markovian \citep{Liow2007,Kiessling2013,Foote2007b,Pigot2012}. In the former case, knowledge of the specific past trajectory of a species -- whether its range has expanded or contracted over a given time span -- might help to improve assessments of its current risk. In the latter case, only the current range of the species conveys useful information about its current risk, although we can still use prior extinction patterns to augment predictions by evaluating the relative extinction risk of species that had similar ranges in the past. Discriminating among these alternative models is thus very important for determining how best to incorporate fossil data in present risk assessments.

@@

We analyzed microfossil occurrence information from the Neptune Database \url{http://www.nsb-mfn-berlin.de/nannotax} \citep{Lazarus1994,SpencerCervato1999}. This occurrence-based dataset includes calcareous nannoplankton, diatoms, foraminifera, and radiolarians.% -- these occurrences span the entire globe between 120 and 0 million years ago (Mya). 
Occurrences were filtered to include only those species with first occurrence was no more than 63 Mya. This filtering criterion excludes taxa that survived the K/Pg extinction or arose during its ecologically unusual recovery interval, and ensures that our occurrence histories fully overlap with the temperature time-series used as a potential extinction risk predictor (see below).

@@

Average global temperature of each 1 My bin was calculated from estimates based on Magnesium/Calcium isotope ratios \citet{Cramer2011}. We use Mg/Ca rather than oxygen isotopes to avoid confounding effect of varying ice-volume -- this property is of particular importance for this analysis as polar ice-caps develop midway through the Cenozoic. Our data source, \citet{Cramer2011}, estimated temperature for every 0.1 My from 0 to 63 Mya. The temperature estimate for each 1 My interval was calculated as the mean of all estimates within that interval.

@@

We used a discrete-time survival modelling framework to estimate how well we can predict extinction risk at one million year time scales. At its core, our model is a multilevel logistic regression with taxon age in millions of years as a varying intercept \citep{Tutz2016}. We considered four different models involving different permutations of covariate effects (fixed or time-varying) and historical covariates: covariate effects constant over time and no historical covariates included (Model C), covariate effects allowed to vary over time but no historical covariates included (Model V), covariate effects constant over time and historical covariates included (Model CP), and covariate effects allowed to vary over time and historical covariates are included (Model VP). The C and P models atempt to predict based only on present state, wheras the CP and VP models allow for the possibility of non-Markovian behaviour by including change in state from the previous time increment.

We always included species age at time of observation (.e. observed prior duration) as a non-nested varying intercept term. This factor may or may not contribute to differences in species extinction risk over time \citep{Smits2015,Finnegan2008,Ezard2012,VanValen1973,Liow2011,Crampton2016}, but its inclusion in our model is critical to its nature as a survival model \cite{Tutz2016}.

@@

We are interested in our models' performance in two contexts: in-sample performance, and out-of-sample predictive performance (i.e. forecasting).

@@

We are particularly interested in understanding how well our model forecasts extinction probability of data from the future that the model was not fit to (out-of-sample data). To quantify our ability to forecast species' extinction risk, we estimated average out-of-sample forecasting performance using 5-fold time-series cross-validation. For time-series data, the folds (data partitions) are approximately equal segments of time. Each fold represents a sequence of time points. With 63 time points, each of the five folds represents approximately 13 million-year time increments. It is important to bear in mind, however, that each time increment includes many (100s-1000s) individual observations.

@@

The in-sample model comparisons are useful for comparing the relative ability of our models to represent the data they were fit to. Comparison between the posterior distributions of in-sample AUC for each of the four models demonstrates that all of our models have approximately equal in-sample forecasting performance (Fig. \ref{fig:auc_hist}). The parameter rich model VP has the greatest median in-sample AUC when compared to the other three models, but there is substantial overlap in their posterior distributions. Additionally, while our parameter rich model VP is possibly the most adequately performing model, the difference or improvement to performance is minimal at best -- all four models have approximately equal in-sample AUC posterior distributions. All the in-sample AUC estimates from our models are concentrated on an AUC value of 0.77. It is therefore hard to conclude that there is one ``best'' model which we can rely upon as they are all nearly functional equivalent.

@@

Depending on the taxon-model combination, there are between zero and 4 time intervals where our posterior distribution of in-sample AUC has a median value less than or equal to 0.5 (Fig. \ref{fig:auc_taxon_time}). However, this pattern is absent for the posterior estimates of in-sample AUC for Foraminfera and Radiolaria as fit by the VP model. In contrast, there are fewer periods of low model performance for calcareous nannoplankton and Dinoflagellates as estimated from our VP model than in those estimates from the other three model variations.

@@

Expected out-of-sample forecasting performance was estimated using five-fold cross-validation for time series \citep{Arlot2010,Bergmeir2016}. The resulting distribution when all folds are combined is highly multimodal as expected given that they are fit to and estimated from different data sets with different numbers of observations \citep{ESL}. This multimodality increases with model complexity (Fig. \ref{fig:fold_auc}), most likely because the complex models allow for predictor effects to vary with time, allowing a greater range in possible parameter values which in turn yield a greater range of posterior predictions.

Comparison the in-sample AUC estimates to the expected out-of-sample AUC estimates reveals a similar range in performance for all models (Fig. \ref{fig:auc_hist}, \ref{fig:fold_auc}). Interestingly, the differences in posterior predictive distributions of AUC between the four model variations is reduced in out-of-sample prediction. For example, the VP model no longer has the greatest median AUC of the four models (Fig. \ref{fig:fold_auc}). For this reason the rank order of median out-of-sample AUC is different from the rank order of median in-sample AUC. However, making interpretations based only on the median AUC estimates is incorrect -- our estimates are the full posterior, not individual points. Because of this, the four models are effectively indistinguishable in their out-of-sample forecast performance (Fig. \ref{fig:fold_auc}).

@@

The posterior predictive distribution of expected out-of-sample AUC over time and across taxonomic groups are nearly identical for each of the four models (Fig. \ref{fig:fold_auc_taxon_time}).

In the analysis of the in-sample forecast performance of the four models, we noted that there were time intervals where our predictions were no better than random (Fig. \ref{fig:auc_taxon_time}). This occurrence is generally much rarer for the posterior distribution of AUC from the out-of-sample forecasts. The major exception to this pattern are our estimates for the Dinoflagellates, which have at least one time interval for all four models in which the median AUC of the out-of-sample forecasts were no better random. The only other group for which median posterior predictive estimate of out-of-sample AUC reaches 0.5 is calcareous nannoplankton, and then only with the V model.
\begin{figure}[ht]
 \centering
 \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_taxon_time_full}
 \caption{Comparison of out-of-sample AUC values over time as aggregated by taxonomic group for each of the four models. The AUC of the individual My intervals within each fold is plotted to highlight the heterogeneity in performance within and between folds. This presentation decomposes each of the 12-million year folds by each of the taxonomic groups into the predictions made for each of the million-year intervals. The black line corresponds to the median AUC estimate, with the envelopes corresponding to multiple credible intervals as indicated in the legend. See Table \ref{tab:model_def} for a description of each of the four models.}
 \label{fig:fold_auc_taxon_time}
\end{figure}

We compared the difference in our AUC estimates from the out-of-sample forecasts to the AUC estimates from our in-sample forecasts by subtracting the in-sample AUC estimates from the out-of-sample AUC estimates (Fig. \ref{fig:oos_ins_diff}). A difference in AUC close to 0 indicates complete congruence between the in-sample and out-of-sample forecasts. A positive difference indicates that our out-of-sample forecasts are actually higher performing than our in-sample forecasts, while negative difference indicates poorer out-of-sample performance than in-sample forecast. Divergences between our out-of-sample and in-sample forecasts are rare and tend to not form multimillion year patters, consistent with the broad visual congruence between the in-sample and out-of-sample forecast performance (Fig. \ref{fig:auc_taxon_time}, \ref{fig:fold_auc_taxon_time}). The only major multimillion year pattern indicating significantly poorer out-of-sample forecast performance than in-sample forecast performance is for Radiolaria based on the VP model concentrated around 30 Mya (Fig. \ref{fig:oos_ins_diff}).

@@

We find that all of our models are expected to correctly forecasting the rank order in extinction probabilities for a randomly selected extinct-extant pair of future observations between 70\% to 80\% of the time (Fig. \ref{fig:fold_auc}). One of the most striking aspects of these results is the similarity in forecasting performance for in-sample and out-of-sample observations. A slight decrease in performance when dealing with out-of-sample observations makes sense: each of the models fit during cross-validation is based on fewer data than the model fit on the full data (between 1/5th to 4/5ths of the original). Additionally, a potential decrease in precision when forecasting the future of extinction risk is to be expected as the future will always differ from the past in some respect. However, the similarly between the in-sample and out-of-sample results indicates that our model is fairly robust to variation in extinction intensity over the Cenozoic. A extremely important caveat, of course, is that human impacts may substantially alter present and future extinction risk dynamics relative to the average Cenozoic state, so that the future may become less predictable than it has been in the past \citep{Harnik2012a,Finnegan2015}.

@@

We also find that our four models are practically identical in their ability to make in-sample and out-of-sample forecasts. Although the in-sample AUC estimates differ between models, all of these estimates are in a narrow range of possible AUC values (Fig. \ref{fig:auc_hist}). While the ``best'' model does include the historical covariates and allows all covariate effects to vary over time, its practical difference in performance versus the other models is negligible (Fig. \ref{fig:fold_auc}). Thus even though the VP model has a statistically greater AUC for in-sample forecasts than the other three models, this result is not practically or scientifically significant. The results of the out-of-sample forecasts illustrate this point in full, as all four models have functionally identical out-of-sample performance. This including geographic range trajectory results in only very minor improvements in forecasting accuracy. This contrasts with findings from coral genera \citep{Kiessling2016} and suggests that further investigation of other taxa, timescales, and environments may help to elucidate the conditions under which past geographic range trajectories are most informative about future extinction risk.

@@

The relative quality and consistency of our models out-of-sample forecasting performance is encouraging given that its estimates are based on very limited biological and environmental information about the taxa. Even our most complex models only account for a few simple aspects of geographic range, prior history, and taxonomy. The principal reason we were not able to include more biological information in our models is that we lack suitably detailed life history or ecological information for many marine micro- and nannoplankton. Foraminifera are an exception to this problem as aspects of life history, ecology, and physiology are known for many foram species \citep{Ezard2011}. However, comparable information does not exist all foram species, nor does this type of data exist for the other three taxonomic groups studied here. Further analyses including this type of information and focused more narrowly on the foraminifera may be informative.

@@

In summary, our results suggest that models trained on prior extinction/survival patterns do modestly well at predicting the relative extinction risk of two randomly selected species based on a small number of taxonomic, geographic, and historical predictors. Considering that conservation decisions are made based on a continuum of risk, from most to least, this means that our results are in the same language as how conservation resources are allocated. These results present an encouraging baseline for using the fossil record to predict future extinction. This suggests, that all else being equal, extending this model to include the information in modern risk assessments might actually do decently well at forecasting future extinctions.
